---
title: "Effects on MPG, Automatic vs. Manual Transmission"
author: "Tyler Richardett"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

Using the ```mtcars``` data set, we answer the following two questions:

1. Is an automatic or manual transmission better for MPG?
2. Quantify the MPG difference between automatic and manual transmissions.

A two-sided t-test concludes that manual transmission is better for MPG. And a multivariate linear regression concludes that manual transmission accounts for a roughly 3 MPG difference over automatic transmission.

## Loading and Cleaning the Data

First, load in the data from the {datasets} library.

``` {r load.data, message = FALSE}
library(datasets)
data(mtcars)
```

Reclassify the "vs" and "am" variables as factors.

```{r reclassify.factors}
mtcars$vs <- factor(mtcars$vs, labels = c("V", "Straight"))
mtcars$am <- factor(mtcars$am, labels = c("Automatic", "Manual"))
```

## Exploratory Data Analysis

Use the ```summary``` function to get a glimpse of the clean data. In the interest of saving space, we'll just look at the "mpg" and "am" variables.

``` {r summary.data}
summary(mtcars[,c(1,9)])
```

See **Appendix A** for a boxplot that compares range of average MPG for vehicles with automatic and manual transmissions.

## Comparing Automatic and Manual Transmissions

Using the ```summarize``` function from the {dplyr} library, calculate the mean MPG for vehicles with automatic and manual transmissions.

```{r calc.means, message = FALSE}
library(dplyr)
am.means <- mtcars %>%
        group_by(am) %>%
        summarize(mean = mean(mpg))
```

``` {r view.means, echo = FALSE, results = "asis", message = FALSE}
library(pander)
pandoc.table(am.means, justify = "left")
```

Because the sample size for each group is low (< 30), use a two-sample t-test to compare the two groups. The null hypothesis we are testing is that there is no difference in MPG between vehicles with automatic transmissions and those with manual transmissions.

``` {r t.test}
t.test(mpg~am, data = mtcars)
```

The p-value (0.001374) is less than 0.05, so we reject the null hypothesis. This answers our first question: **Manual transmission is better than automatic transmission for MPG**.

## Quantifying the Difference Between Automatic and Manual Transmissions

Use the ```step``` function to choose the best model.

``` {r step.model, results = "hide"}
step.model <- step(lm(mpg ~ ., data = mtcars))
```

``` {r summary.step.model, echo = FALSE, results = "as.is"}
summary(step.model)
```

As you can see, the best fit model includes the "am" variable, which is fortunate for the purposes of this analysis, and it explains about 85 percent of the variance. See **Appendix B** for the residual plots, which show no clear pattern.

This answers our second question: **Manual transmission improves MPG by 2.94 units, over automatic transmission**. The p-value (0.046716) is less than 0.05.

## Appendices

### Appendix A

Use ```ggplot``` to create a boxplot of MPG by automatic vs. manual transmissions.

``` {r boxplot, message = FALSE}
library(ggplot2)
ggplot(mtcars, aes(x = am, y = mpg)) +
        geom_boxplot() +
        labs(y = "Miles per Gallon (MPG)", x = "Transmission Type") +
        ggtitle("MPG Distribution, Automatic vs. Manual")
```

\newpage
And use ```ggpairs``` to create a matrix that shows the intersection of all variables.

``` {r ggpairs, message = FALSE, cache = TRUE}
library(GGally)
ggpairs(mtcars)
```

\newpage
### Appendix B

Below, find the residual plots for the model used in the analysis.

``` {r plot.resids}
par(mfrow = c(2,2))
plot(step.model)
```